% ICASSP submission: spconf.sty + IEEEbib.bst (official paper-kit files).
% At most 4 pages of technical content; page 5 contains ONLY references.
\documentclass{article}
\usepackage{spconf,amsmath,amssymb,amsthm,graphicx,bm,booktabs}
\ninept
\graphicspath{{figures/}}

\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{remark}{Remark}

\emergencystretch=1em
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\tr}{tr}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\W}{\mathcal{W}}
\newcommand{\Rlb}{R_{\mathrm{LB}}}
\newcommand{\norm}[1]{\lVert#1\rVert}

% Real-audio numbers are injected from experiment JSON (scripts/make_paper_macros.py).
\IfFileExists{macros_audio.tex}{\input{macros_audio.tex}}{%
  \newcommand{\nTest}{[pending]}
  \newcommand{\nCalib}{[pending]}
  \newcommand{\nSpeakersTest}{[pending]}
  \newcommand{\pesqTargetVal}{[pending]}
  \newcommand{\spearmanSens}{[pending]}
  \newcommand{\spearmanSensCI}{[pending]}
  \newcommand{\rTwoSens}{[pending]}
  \newcommand{\permP}{[pending]}
  \newcommand{\spearmanSNR}{[pending]}
  \newcommand{\spearmanCentroid}{[pending]}
  \newcommand{\nullLeakEps}{[pending]}
  \newcommand{\nullLeakOp}{[pending]}
  \newcommand{\rowLeakOp}{[pending]}
}

\title{A Subspace Model of Speech-Watermark Survival\\ Under Analysis--Resynthesis}
\name{Weiping Yan\thanks{Code, proofs, per-clip score arrays, and locked environments:
  \texttt{github.com/\allowbreak appleweiping/\allowbreak resynthesis-watermark-limits}.}}
\address{University of Minnesota, Twin Cities, Minneapolis, MN, USA}

\begin{document}
\maketitle

\begin{abstract}
Speech watermarks are widely reported to fail when audio is regenerated by a vocoder or
neural codec. We give this phenomenon a restricted but precise model: a blind launderer is
an \emph{analysis--resynthesis channel} $\W=\mathcal S\circ\A$. At full generality we prove
only what is true: a data-processing inequality (laundering cannot increase the Chernoff
detection exponent), and an \emph{exact-erasure condition} --- if embedding leaves the
analysis output unchanged a.s., the laundered laws coincide and no detector beats chance.
In a linear-Gaussian surrogate we obtain closed forms: the channel transmits exactly the
equivalence class of the perturbation modulo $\ker\A$, with surviving exponent
$\tfrac18\norm{P\delta}^2$; a Gaussian \emph{achievable lower bound} $\Rlb$ (not a
capacity) covers blind embedders. The model's empirical content is a measurable,
channel-relative \emph{survival predictor}: each attacker's own analysis sensitivity.
On LibriSpeech ($\nTest$ speaker-disjoint test clips, thresholds from $\nCalib$ independent
calibration clips), across spectral inversion, two neural vocoder pipelines, three neural
codecs, and self voice-conversion, the predictor tracks held-out post-attack separability
(Spearman $\spearmanSens$, perm.\ $p=\permP$), outperforming SNR and spectral-centroid
controls. Deployed watermarks (AudioSeal, WavMark, SilentCipher) at matched perceptual
quality fail \emph{operationally} under resynthesis --- score sign/order inversions under
their original operating rules --- while attack-conditioned separability often remains;
we report both, and never conflate them.
\end{abstract}

\begin{keywords}
audio watermarking, generative AI, provenance, data-processing inequality, hypothesis testing
\end{keywords}

\section{Introduction}
\label{sec:intro}
Provenance watermarking embeds a key-dependent signal in speech so a detector can later
flag synthetic or traced audio~\cite{sanroman2024audioseal,chen2023wavmark,
singh2024silentcipher}. A consistent empirical finding is fragility to
\emph{regeneration}: passing watermarked audio through a vocoder, neural codec, or voice
conversion drives deployed detectors toward failure~\cite{oreilly2025shallow,wen2025sok}.
The community's response is to embed in representations that resynthesis preserves ---
codec latents (Latent-Mark~\cite{chen2026latentmark}), speech-feature-aligned
marks~\cite{li2026featurealigned}, speaker-latents robust to zero-shot
VC~\cite{li2025voicemark}, and semantic image watermarks before
them~\cite{zhao2024invisible,wen2023treering,yang2024gaussianshading}. The
\emph{invariant-alignment idea is thus established practice, not our contribution.}

What the literature lacks is a stated model in which the collapse and its exceptions can
be derived and \emph{predicted}. Empirical papers report which schemes die under which
attacks; the pattern --- AudioSeal survives EnCodec yet fails under mel-inversion; marks
with more energy in preserved components survive longer --- is left as observation. This
paper supplies the missing object: a channel model with exactly-scoped guarantees, and a
\emph{measurable} quantity that predicts survival out-of-sample.

\noindent\textbf{Contributions.}
(i)~A general proposition for blind analysis--resynthesis channels: a Chernoff
data-processing inequality, and a sufficient condition for \emph{exact} erasure
(\S\ref{sec:model}). Nothing stronger is claimed at this generality.
(ii)~A linear-Gaussian surrogate where everything is closed-form: the channel transmits
the perturbation's class in $\R^n/\ker\A$; mixed perturbations survive \emph{partially}
with exponent $\tfrac18\norm{P\delta}^2$ (\S\ref{sec:lg}).
(iii)~A Gaussian achievable lower bound $\Rlb$ for blind embedders --- explicitly
\emph{not} a watermarking capacity: informed embedding~\cite{costa1983writing,
moulin2003information} can exceed it and no rate converse is claimed (\S\ref{sec:lg}).
(iv)~A channel-relative survival predictor --- each attacker's own analysis
sensitivity --- validated on \emph{held-out, speaker-disjoint} data across seven
resynthesis families against competing predictors, with permutation controls
(\S\ref{sec:exp}).
(v)~A calibrated operational evaluation of deployed watermarks at matched perceptual
quality, separating \emph{operational failure} (sign/order-inverted scores under the
original rule, thresholds from an independent calibration set) from loss of
\emph{separability} (oriented AUC) --- the two are routinely conflated
(\S\ref{sec:exp}).

\section{The analysis--resynthesis channel}
\label{sec:model}
A watermark embedder $E_\kappa$ maps a host $x\sim P_0$ to $E_\kappa(x)$ (additively,
$E_\kappa(x)=x+\delta_\kappa$), giving the watermarked law $P_1$. A detector knowing
$\kappa$ tests $H_0\!:P_0$ vs $H_1\!:P_1$; with $N$ i.i.d.\ looks the optimal Bayes error
decays as $e^{-N\,C(P_0,P_1)}$, $C$ the Chernoff
information~\cite{chernoff1952measure,cover2006elements}. A \emph{blind launderer} is
\begin{equation}
\W \;=\; \mathcal S\circ \A,
\end{equation}
where $\A$ maps a waveform to a representation $z$ (a mel envelope, a codec latent) and
$\mathcal S$ regenerates a waveform from $z$ with randomness independent of the input.
The attacker knows neither $\kappa$ nor the scheme (adaptive, key-aware removal
\cite{ding2026evade} is out of scope). Imperceptibility is a budget $\rho(\delta;x)\le D$.

\begin{proposition}[general channels]\label{prop:general}
For any channel $\W$,
$C(\W_\#P_0,\W_\#P_1)\le C(P_0,P_1)$.
If moreover $\A(E_\kappa(x))=\A(x)$ holds $P_0$-a.s.\ and $\mathcal S$ depends on its
input only through $z$, then $\W_\#P_1=\W_\#P_0$; hence
$C(\W_\#P_0,\W_\#P_1)=0$ and no detector beats chance at any sample size.
\end{proposition}

\begin{proof}
The first part is the DPI for the Chernoff family: each
$-\log\int p_0^{1-s}p_1^s$ is monotone under Markov kernels and the max over $s$
preserves monotonicity~\cite{cover2006elements}. For the second,
$\W(E_\kappa(x))=\mathcal S(\A(E_\kappa(x)),U)=\mathcal S(\A(x),U)=\W(x)$ a.s.\ with
$U\perp x$, so the output laws coincide.
\end{proof}

Proposition~\ref{prop:general} is all we assert for real nonlinear vocoders and codecs:
laundering cannot help the detector, and \emph{exactly} analysis-invariant embeddings are
\emph{exactly} erased. It supplies no constants and no prediction of partial survival.
Quantitative structure requires a model.

\section{Linear-Gaussian surrogate}
\label{sec:lg}
Work in whitened coordinates, $P_0=\N(0,I_n)$, perceptual weighting carried by a masking
metric $M\succ0$ ($\rho(\delta)=\delta^\top M\delta$). Let $\A\in\R^{k\times n}$ have full
row rank, $P=\A^{+}\A$ the orthogonal projector onto $\mathrm{row}(\A)$, and let synthesis
resample the complement from the clean prior:
\begin{equation}\label{eq:lgchannel}
\W(x)=Px+(I-P)w,\qquad w\sim\N(0,I_n)\ \text{fresh}.
\end{equation}

\begin{theorem}[exact surviving shift and exponent]\label{thm:lg}
Under \eqref{eq:lgchannel}, for a fixed additive shift $\delta$:
$\W_\#\N(\delta,I)=\N(P\delta,I)$ and
\begin{equation}
C(\W_\#P_0,\W_\#P_1)\;=\;\tfrac18\norm{P\delta}^2 .
\end{equation}
\end{theorem}

The channel transmits exactly the equivalence class $[\delta]\in\R^n/\ker\A$: the
effective perturbation is $P\delta$. Three consequences, stated carefully:
$\delta\in\ker\A$ is erased exactly (recovering Prop.~\ref{prop:general});
\emph{mixed} perturbations survive \emph{partially} (exponent
$\tfrac18\norm{P\delta}^2>0$ whenever $P\delta\neq0$); and row-space-only perturbations
waste no budget on erased components but are \emph{not} the unique surviving set. The
budget-optimal exponent is $\max_{\delta^\top M\delta\le D}\tfrac18\norm{P\delta}^2 =
\tfrac D8\,\lambda_{\max}(P,M)$, at the top generalized eigenvector of $(P,M)$.

\noindent\textbf{A rate lower bound (not capacity).} Writing surviving shifts in an
orthonormal basis of $\mathrm{row}(\A)$, a \emph{blind} (host-uninformed, Gaussian
codebook) embedder facing a $\kappa$-aware but host-blind decoder achieves any rate below
\begin{equation}\label{eq:rlb}
\Rlb=\max_{Q\succeq0,\ \tr(M_{\mathrm{row}}Q)\le D}\tfrac12\log\det(I+Q)
\end{equation}
nats per invariant coordinate (host as unit-variance interference; water-filling over
$M_{\mathrm{row}}$'s eigenmodes). $\Rlb$ is an \emph{achievable lower bound for that
restricted class}: informed (host-aware) embedding in the Costa/Moulin--O'Sullivan
sense~\cite{costa1983writing,moulin2003information} can exceed it, and we prove no rate
converse. A nullspace embedding has $\Rlb=0$ after $\W$.

\noindent\textbf{From surrogate to real channels.} For a real attacker $\W$ the model
suggests a \emph{measurable} channel-relative predictor: the attacker's own analysis
sensitivity
\begin{equation}\label{eq:sens}
s_{\W}(\delta;x)=\norm{\A_{\W}(x+\delta)-\A_{\W}(x)}/\norm{\delta},
\end{equation}
computed with \emph{that} attacker's $\A_\W$ (a mel front-end, a codec encoder). Whether
$s_\W$ predicts post-attack separability is an \emph{empirical hypothesis} --- the
surrogate motivates it but does not prove it. Testing it out-of-sample is E2's job; a
single mel fraction is never reused across attackers with different $\A_\W$.

\section{Experiments}
\label{sec:exp}
\noindent\textbf{Setup.} LibriSpeech 16\,kHz: test $=$ \nTest{} speaker-stratified
4\,s voice-active clips from test-clean (\nSpeakersTest{} speakers); detector thresholds
come \emph{only} from \nCalib{} independent calibration clips from dev-clean
(speaker-disjoint from test by partition); the predictor's mapping is fit on a dev-clean
fit split and evaluated on test-clean only. Attackers: STFT-GL (near-lossless control),
mel-GL (spectral inversion of THE unified 80-mel analysis), Vocos (neural vocoder),
Vocos-EnCodec (learned reconstruction from codec tokens), EnCodec at $6/3/1.5$\,kbps,
DAC, SNAC, and kNN-VC self voice-conversion (top-$k\in\{4,8\}$) --- content-, speaker-
and prosody-preserving re-synthesis. All formal runs fail loudly if any attacker or
baseline is missing. Scores, seeds, keys, manifests, and locked environments ship with
the repository.

\noindent\textbf{Constructed marks (geometry probes).} W.r.t.\ the unified mel analysis
$\A$ we construct, per utterance and key: a \emph{kernel mark} (alternating projections
onto quadrature $\cap$ consistent spectrograms --- the first-order kernel of magnitude
analysis) and a \emph{row mark} (in-phase mel-pattern shifts). Construction quality is
\emph{measured}, not assumed: at infinitesimal amplitude the kernel mark's analysis
change is $\nullLeakEps\times$ a random direction's; at the operating amplitude the
(second-order) leakage is $\nullLeakOp\times$ random, and the row mark's is
$\rowLeakOp\times$. Their detection uses a genie-aided matched-filter \emph{probe}
(the known per-utterance perturbation) --- a diagnostic of channel geometry, explicitly
not a deployable detector; deployed-detector conclusions come only from the baselines.

\noindent\textbf{E2 --- does channel-relative sensitivity predict survival?}
For $\ge$100 random directions (kernel/row mixtures, random in-phase) $\times$
randomized budgets, we measure $s_\W$ \eqref{eq:sens} per attacker and the oriented
post-attack AUC of the probe. The survival mapping (isotonic, and
$\Phi(a\sqrt f{+}b)$) is fit on dev-clean and evaluated on test-clean:
Spearman $\spearmanSens$ (95\% CI $\spearmanSensCI$), $R^2=\rTwoSens$;
within-attacker permutation test $p=\permP$. Competitors under the identical protocol:
waveform SNR ($\rho_s=\spearmanSNR$), spectral centroid ($\rho_s=\spearmanCentroid$),
STFT-magnitude fraction. [Table/figure inserted from results.]

\noindent\textbf{E1 --- deployed watermarks at matched perceptual quality.}
AudioSeal~\cite{sanroman2024audioseal}, WavMark~\cite{chen2023wavmark}, and
SilentCipher~\cite{singh2024silentcipher}, each tuned to the same median PESQ
(\pesqTargetVal) before attack --- no audible-vs-transparent comparisons. Per
(baseline, attacker, key): raw AUC, oriented AUC (utterance-clustered bootstrap CIs),
TPR at the calibration-fixed 1\%-FPR threshold (Clopper--Pearson), payload accuracy, and
an attack-aware recalibration diagnostic. [Table inserted from results.] The recurring
pattern: under resynthesis the deployed scores often \emph{invert} (raw AUC $<0.5$) ---
operational failure under the original rule --- while oriented separability remains
substantial; codec-hardened schemes survive the codecs they were hardened against. This
is precisely the detector-dependence the subspace model predicts: survival tracks where
the \emph{detector reads}, relative to what the channel preserves.

\noindent\textbf{E3 --- multi-bit proof of concept (no rate claim).}
A $B$-bit payload in the unified mel domain with frame-differential BPSK and a fully
\emph{blind} decoder survives mel-GL, Vocos, and codec round-trips with BER/BLER
reported with binomial CIs. This demonstrates the invariant sub-channel carries bits; it
is \emph{not} an achievable-rate measurement and we claim no bits/s at target
reliability.

\section{Relation to prior work}
Attack studies document the collapse~\cite{oreilly2025shallow,wen2025sok}; invariant/
latent-aligned embedding is established practice: Latent-Mark optimizes waveforms to
shift codec latents with cross-codec transfer~\cite{chen2026latentmark}, Feature-Aligned
marks match speech-feature distributions~\cite{li2026featurealigned}, VoiceMark targets
VC-invariant speaker latents~\cite{li2025voicemark}; in images, semantic watermarks
realize the same prescription~\cite{zhao2024invisible,wen2023treering,
yang2024gaussianshading}. Relative to all of these we add what a method paper does not:
a stated channel model with exactly-scoped guarantees (Prop.~\ref{prop:general},
Thm.~\ref{thm:lg}), an explicit account of what is \emph{not} proven ($\Rlb$ vs
capacity~\cite{moulin2003information,costa1983writing,chen2001quantization}), and a
falsifiable, channel-relative predictor validated out-of-sample with permutation
controls. Our evaluation protocol (independent calibration, oriented-vs-raw separation)
addresses measurement pathologies that affect published robustness numbers.

\section{Scope and limitations}
\label{sec:disc}
The theory covers blind, non-adaptive laundering; key-aware
removal~\cite{ding2026evade} is out of scope. Exact erasure requires exact analysis
invariance --- our constructed kernel marks realize it only to first order, with
measured second-order leakage at operating amplitude. $\Rlb$ concerns a restricted
embedder class; real capacities may be higher. The surrogate's closed forms are not
claimed for nonlinear generators --- for those we claim only Prop.~\ref{prop:general}
and the \emph{measured} predictive power of $s_\W$. Survival statements are about
oriented separability or calibrated operating points, never raw sub-0.5 AUC read as
erasure.

\section{Conclusion}
A restricted but precise subspace model of analysis--resynthesis explains, with
correctly-scoped guarantees, when speech watermarks survive laundering, and yields a
measurable channel-relative predictor that generalizes across attackers and held-out
speakers. Honest accounting --- of what is proved, what is measured, and what deployed
detectors actually do under attack --- is the paper's method as much as its message.

\vfill\pagebreak

\bibliographystyle{IEEEbib}
\bibliography{refs}

\end{document}
